Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2014-07-18 pituba/article/article_spe.tex
misc add

The material balance equation is widely used to relate production, injection and reservoir pressure. Actually, all averaged fluid properties in Eq.~\eqref{eq: MBE} depend on the pressure field, and so does the pore volume. If the reservoir is under production, though, the pressure field is not known and for its prediction one would have to resort to more advanced techniques such as reservoir simulation.

@@

To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume hydrostatic equilibrium: $\vec{u}_\alpha \equiv 0$. This implies that $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, as a consequence, pressure becomes a function of time and depth only given by the following expression:
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha^{ref} (t)}^{p_\alpha (t,z)} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'}\right)} = g\left(z - z_{ref}\right) \, ,
\end{equation}
where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha^{ref} (t)$ is its pressure at some reference depth $z_{ref}$.
It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly.

Furthermore, it can be shown that neglecting the variation of fluid properties with depth does not introduce any important error. As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. For instance, the average oil density can be approximated to the local density at the reference level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$; the same applies to the formation volume factor: $B_o \approx \rho_o (t, z_{ref})/ \rho_o^{ST}$.

@@

For investigating the order of magnitude of Eqs.~\eqref{eq: gradodz} and \eqref{eq: gradgdz}, we substitute in them the values of $\rho_o = 10^3\ kg/m^3$ and $c_o = 10^{-9}\ Pa^{-1}$ for the oil phase, and $\mathcal{M}=19\ kg/kmol$ and $T=400\ K$ for the gas phase: 
\begin{equation}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z}\right)_{T,Y_o} \approx 1\times10^{-5}\ m^{-1} \, ,
\end{equation}

@@

Oil and gas densities are thus approximately constant with depth.
The reference depth for computing phase density (and also the formation volume factor) can be any arbitrary depth inside phase volume. Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$, where $z^0_\alpha$ is the depth of phase gravity center and $\alpha=w, o, g$. The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$, without fluid motion, and only equals the average reservoir pressure $\bar{p}_\alpha$ if the reference depth is the depth of the gravity center, $z^{ref}=z^0$.

@@

Residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except for the first order phase change at bubble-point pressure.
The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels, each corresponding to a different phase (three fluid phases and the rock). For the fluid phases, each parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
\begin{equation}
\begin{split}
&V_{p1}\frac{d\varepsilon}{dp}=\underbrace{\left(N-N_p\right) \left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right)}_\text{oil phase}\\
&+\underbrace{\left[N R_{s1}+mN\frac{B_{o1}}{B_{g1}}+G_i-\left(N-N_p\right)R_{s2}-G_p\right] \frac{dB_g}{dp}}_\text{gas phase}\\
&+\underbrace{\left(W + W_i+W_e-W_p\right)\frac{dB_w}{dp}}_\text{water phase}\\
&\underbrace{-\frac{dV_p}{dp}}_\text{porous volume}\, .
\end{split}
\end{equation}

>>>>

The material balance equation is widely used to relate production and injection to reservoir pressure. Actually, all averaged fluid properties in Eq.~\eqref{eq: MBE} depend on the pressure field, and so does the pore volume. If the reservoir is under production, though, the pressure field is not known and for its prediction one would have to resort to more advanced techniques such as reservoir simulation.

@@

To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume hydrostatic equilibrium: $\vec{u}_\alpha \equiv 0$. This implies that $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, as a consequence, pressure becomes a function of time and depth only given by the following expression:
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha^{ref} (t)}^{p_\alpha (t,z)} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'}\right)} = g\left(z - z_{ref}\right) \, ,
\end{equation}
where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha^{ref} (t)$ is the pressure at some reference depth $z_{ref}$.
It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly.

It can be shown that neglecting the variation of fluid properties with depth does not introduce any important error. As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. For instance, the average oil density can be approximated to the local density at the reference level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$; the same applies to the formation volume factor: $B_o \approx \rho_o (t, z_{ref})/ \rho_o^{ST}$.

@@

For investigating the order of magnitude of Eqs.~\eqref{eq: gradodz} and \eqref{eq: gradgdz}, we substitute into them the values of $\rho_o = 10^3\ kg/m^3$ and $c_o = 10^{-9}\ Pa^{-1}$ for the oil phase, and $\mathcal{M}=19\ kg/kmol$ and $T=400\ K$ for the gas phase: 
\begin{equation}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z}\right)_{T,Y_o} \approx 1\times10^{-5}\ m^{-1} \, ,
\end{equation}

@@

Oil and gas densities are thus approximately constant with depth.
The reference depth for computing phase density (and also the formation volume factor) can be any arbitrary depth inside phase volume. Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$, where $z^0_\alpha$ is the depth of phase gravity center and $\alpha=w, o, g$. The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$, without fluid motion, and only equals the average reservoir pressure $\bar{p}_\alpha$ if the reference depth is that of the gravity center, $z^{ref}=z^0$.

@@

The residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except for the first order phase change at bubble-point pressure.
The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels, each corresponding to a different phase (three fluid phases and the solid rock). For the fluid phases, each parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
\begin{equation}
\begin{split}
&V_{p1}\frac{d\varepsilon}{dp}=\underbrace{\left(N-N_p\right) \left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right)}_\text{oil phase}\\
&+\underbrace{\left[N R_{s1}+mN\frac{B_{o1}}{B_{g1}}+G_i-\left(N-N_p\right)R_{s2}-G_p\right] \frac{dB_g}{dp}}_\text{gas phase}\\
&+\underbrace{\left(W + W_i+W_e-W_p\right)\frac{dB_w}{dp}}_\text{water phase}\\
&\underbrace{-\frac{dV_p}{dp}}_\text{porous volume}\, .
\end{split}
\end{equation}
